### 2021-04-12 paper/ase21/analyzer.tex
[ASE21] Figure 5 - abstract semantics

$\ires$ is an untyped intermediate representation introduced by \citet{jiset}.
To make it more suitable for type analysis, we modify it as a label-based
language as follows:
\[
  \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
    \text{Functions}
    &\funcset&\ni&\func&::=&\kwdef \; \x (\x^*, \x^*). \; \lab\\

    \text{Instructions}
    &\instset&\ni&\inst&::=&
    \kwlet \; \x = \expr \mid
    \kwlet \; \x = \kwrl \expr \; \expr^* \kwrr \mid
    \kwassert \; \expr \\

@@

    \text{Types}
    &\tyset&\ni&\ty&::=&\tname \mid \tjs \mid \tprim \mid
    \undefval \mid \nullval \mid \tbool\\

    &&&&\mid&
    \tnumeric \mid \tnum \mid \tbigint \mid \tstr \mid \tsymb \\
  \end{array}
\]
A modified $\ires$ program $\prog = (\funcs, \getlab, \nextlab)$ is a triple of
a set of functions $\funcs \subseteq \and \funcset$ and two finite mappings
$\getlab$ and $nextlab$.  The getter mapping $\getlab: \labset \finmap \instset$
represents instructions attached to labels, and the next mapping $\nextlab:
\labset \finmap \labset$ denotes next labels.  A function $\kwdef \; \f (\x^*,
\y^*). \; \lab \in \funcset$ consists of a name $\f$, normal parameters $\x^*$,
optional parameters $\y^*$, and a body label $\lab$.  A label $\lab \in \labset$
denotes a program point.  An instruction $\inst$ is a variable declaration, a
function call, an assertion, a branch, a return, or an reference update.  Each
invocation of an abstract algorithm in ECMAScript is compiled to a function call
instruction with a newly introduced temporal variable.  We represent loops using
branch instructions with cyclic pointing of labels in the next mapping
$\nextlab$.  A reference $\refer$ is a variable $\x$ or a field access $\refer
\kwsl \expr \kwsr$.  An expression $\expr$ is a record, a list, a type check, an
existence check, a binary operation, a unary operation, a reference, a constant,
or a primitive, which is either $\undefval$, $\nullval$, a Boolean $\bool$, a
Number $\num$, a BigInt $\bigint$, a String $\str$, or a Symbol $\symb$.

\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/subtype}
  \vspace*{-1.5em}
  \caption{A graphical representation of subtype relation $\subtype$.}
  \label{fig:subtype}
  \vspace*{-1.5em}
\end{figure}

A type $\ty \in \tyset$ is a norminal type $\tname$ or a pre-defined type;
$\tjs$ denotes JavaScript values, $\tprim$ primitives, $\tnumeric$ either Number
or BigInt values, and $\tnum$, $\tbigint$, $\tstr$, and $\tsymb$ denote Number,
BigInt, String, and Symbol values, respectively.  A norminal type $\tname$ is a
production name for ASTs or a record type name used in abstract algorithms.  The
subtype relation $\subtype \subseteq \tyset \times \tyset$ between types is
described in Figure~\ref{fig:subtype}; the directed edge from $\ty'$ to $\ty$
denotes $\ty' \subtype \ty$ and the relation is reflexive and transitive.  The
subtype relation is dependent on the syntax and defined record types in
ECMAScript.  We manually model subtypes between record types but automatically
extract subtypes for AST types.  For example, consider the syntax-directed
abstract algorithm at the right bottom in Figure~\ref{fig:example}.  The
non-terminal \textit{BindingElement} is the unique token in a alternative of
\textit{FormalParameter} thus we automatically extract the subtype relation:
\textit{BindingElement} $\subtype$ \textit{FormalParameter}.  Based on this
subtype relation, a type check expression $\expr: \ty$ checks whether the
evaluation result of $\expr$ has a type $\ty'$ satisfying $\ty' \subtype \ty$.
We represent semantics of the modified $\ires$ as a state transition system
$(\stset, \trans, \istset)$ with the transition relation $\trans \subseteq
\stset \times \stset$.  For presentation brevity, we omit its detail in this
paper and include it in a companion report~\inred{\cite{report}}.
% \[
%   \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
%     \text{States}
%     &\st&\in&\stset&::=&\ctxtset^* \times \labset \times \envset\\
% 
%     \text{Contexts}
%     &\ctxt$\in$\ctxtset&::=&\labset \times \varset \times \envset\\
% 
%     \text{Environments}
%     &\env$\in$\envset&::=&\varset \finmap \valset\\
% 
%     \text{Values}
%     &\val$\in$\valset&::=& \addrset \uplus \addr \uplus \primset\\\
%   \end{array}
% \]

@@

Abstract Interpretation~\cite{ai1977, ai1992}

\begin{figure*}[t]
  \centering
  \[
    \todo
  \]
  \vspace*{-1em}
  \caption{Abstract .}
  \label{fig:example}
\end{figure*}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/record-fields-table}
  \vspace*{-1.5em}
  \caption{Fields of completion records in ES11.}
  \label{fig:record-fields-table}
  \vspace*{-1.5em}
\end{figure}

$\ires$ is an untyped intermediate representation introduced by \citet{jiset}.
To make it more suitable for type analysis, we modify it as a label-based
language as follows:
\small
\[
  \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
    \text{Functions}
    &\funcset&\ni&\func&::=&\kwdef \; \x (\x^*, \x^*). \; \lab\\

    \text{Instructions}
    &\instset&\ni&\inst&::=&
    \kwlet \; \x = \expr \mid
    \x = \kwrl \expr \; \expr^* \kwrr \mid
    \kwassert \; \expr \\

@@

    \text{Types}
    &\tyset&\ni&\ty&::=&\tname \mid \clist{} \mid \clist{\ty} \mid \tjs \mid
    \tprim \mid \undefval \mid \nullval\\

    &&&&\mid&
    \tbool \mid \tnumeric \mid \tnum \mid \tbigint \mid \tstr \mid \tsymb \\
  \end{array}
\]

A modified $\ires$ program $\prog = (\getfunc, \getlab, \getnext)$ is consists
of three mappings from labels;  $\getfunc: \labset \rightarrow \funcset$
represents that labels are in which functions, $\getlab: \labset \rightarrow
\instset$ represents instructions attached to labels, and $\getnext: \labset
\rightarrow \labset$ denotes next labels where a label $\lab \in \labset$
denotes a program point.  function $\kwdef \; \f (\x^*, \y^*). \; \lab \in
\funcset$ consists of its name $\f$, normal parameters $\x^*$, optional
parameters $\y^*$, and a body label $\lab$.  Although global variables exist in
ECMAScript, we assume that only local variables exist in this paper for brevity
An instruction $\inst$ is a variable declaration, a function call, an assertion,
a branch, a return, or an reference update.  Each invocation of an abstract
algorithm in ECMAScript is compiled to a function call instruction with a newly
introduced temporal variable.  We represent loops using branch instructions with
cyclic pointing of labels in the next mapping $\getnext$.  A reference $\refer$
is a variable $\x$ or a field access $\refer \kwsl \expr \kwsr$.  An expression
$\expr$ is a record, a list, a type check, an existence check, a binary
operation, a unary operation, a reference, a constant, or a primitive, which is
either $\undefval$, $\nullval$, a Boolean $\bool$, a Number $\num$, a BigInt
$\bigint$, a String $\str$, or a Symbol $\symb$.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{img/subtype}
  \vspace*{-1.5em}
  \caption{A graphical representation of subtype relation $\subtype$.}
  \label{fig:subtype}
  \vspace*{-1.5em}
\end{figure}

A type $\ty \in \tyset$ is a norminal type $\tname$ or a pre-defined type;
$\tjs$ denotes JavaScript values, $\tprim$ primitives, $\tnumeric$ either Number
or BigInt values, and $\tnum$, $\tbigint$, $\tstr$, and $\tsymb$ denote Number,
BigInt, String, and Symbol values, respectively.  A norminal type $\tname$ is
either 1) a \textit{production type} for ASTs or 2) a \textit{record type} used
in abstract algorithms, and ECMAScript describes its fields with pre-defined or
possible values.  A production type has corresponding syntax-directed algorithms
as its fields and a record type has specific fields with possible values
described in the corresponding table and additional description in ECMAScript.
The production type is automatically extracted but we manually model fields of
record types.  For example, Figure~\label{fig:record-fields-table} is an excerpt
from ES11 that describes fields of completion
records~\footnote{https://262.ecma-international.org/11.0/\#table-8}; it
contains three fields ``Type'' with a constant such as \code{normal}, ``Value''
with JavaScript values or an constant \code{empty}, and ``Target'' with String
values or \code{empty}.  We model them as follows:
\small
\[
  \begin{array}{l}
    \text{Completion} = \kwcl\\
    \quad \text{Type}: \kwcl \nconst{normal}, \nconst{break}, \nconst{continue},
    \nconst{return}, \nconst{throw} \kwcr\\
    \quad \text{Value}: \kwcl \tjs, \nconst{empty} \kwcr\\
    \quad \text{Target}: \kwcl \tstr, \nconst{empty} \kwcr\\
    \kwcr\\
  \end{array}
\]

The subtype relation $\subtype \subseteq \tyset \times \tyset$ between types is
described in Figure~\ref{fig:subtype}; the directed edge from $\ty'$ to $\ty$
denotes $\ty' \subtype \ty$ and the relation is reflexive and transitive.  The
subtype relation is dependent on the syntax and defined record types in
ECMAScript.  We extract subtypes for production types and manually model
subtypes between record types.  For example, consider the syntax-directed
abstract algorithm at the right bottom in Figure~\ref{fig:example}.  The
non-terminal \textit{BindingElement} is the unique token in a alternative of
\textit{FormalParameter} production thus we automatically extract the subtype
relation: \textit{BindingElement} $\subtype$ \textit{FormalParameter}.  Based on
this subtype relation, a type check expression $\expr: \ty$ checks whether the
evaluation result of $\expr$ has a type $\ty'$ satisfying $\ty' \subtype \ty$.

We define a denotational semantics of the modified $\ires$ for instructions
$\sem{\inst}: \dom \rightarrow \dom$ and expressions $\sem{\expr}: \dom
\rightarrow \valset$.  For presentation brevity, we omit it in this paper and
include it in a companion report~\inred{\cite{report}}.
% \[
%   \begin{array}{l@{~}r@{~}c@{~}r@{~}r@{~}l}
%     \text{States}
%     &\elem&\in&\dom&::=&\ctxtset^* \times \labset \times \envset\\
% 
%     \text{Contexts}
%     &\ctxt$\in$\ctxtset&::=&\labset \times \varset \times \envset\\
% 
%     \text{Environments}
%     &\env$\in$\envset&::=&\varset \finmap \valset\\
% 
%     \text{Values}
%     &\val$\in$\valset&::=& \addrset \uplus \addr \uplus \primset\\\
%   \end{array}
% \]

@@

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{\textwidth}
    \[
      \begin{array}{r@{~}c@{~}ll}
        \asemi{\kwlet \; \x = \expr}
        (\lab, \tys, \aenv, \_) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv_\x \}, \emp) &
        \text{where} \;
        \aenv_\x = \aenv[\x \mapsto \aty] \wedge
        \aty = \aseme{\expr}(\aenv)\\

        \asemi{\x = \kwrl \expr_0 \; \expr_1 \cdots \expr_n \kwrr}
        (\lab, \tys, \aenv, \_) &=&
        \left(
          \begin{array}{@{}l@{}}
            \{ (\lab_\func, \tys_a) \mapsto \aenv_{\func, \tys_a} \mid \func \in
              F \wedge \tys_a \in T_a
            \},\\
            \{ (\func, \tys_a) \mapsto \{ (\getnext(\lab), \tys, \x) \} \mid
              \func \in F \wedge \tys_a \in T_a
            \}\\
          \end{array}
        \right) &
        \text{where} \;
        \left\{
          \begin{array}{@{}l@{}}
            \aty_i = \aseme{\expr_i}(\aenv) \wedge T_a = \{ \upcasts([ \ty_1,
            \cdots, \ty_n ] \mid \ty_i \in \aty_i \} \wedge\\
            F = \{ \func \in \aty_0 \mid \func = \kwdef \; \_ (\p_1, \cdots,
            \p_m). \; \lab_\func \} \wedge\\
            \aenv_{\func, \tys_a} = \aenv[\p_1 \mapsto \{ \tys_a[1] \}, \cdots,
            \p_m \mapsto \{ \tys_a[m] \}]\\
          \end{array}
        \right.\\

        \asemi{\kwassert \; \expr}
        (\lab, \tys, \aenv, \_) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv_\mt \}, \emp) &
        \text{where} \; \aenv_\mt = \refine(\aenv, \expr, \true)\\

        \asemi{\kwif \; \expr \; \lab_\mt \; \lab_\mf}
        (\lab, \tys, \aenv, \_) &=&
        (\{ (\lab_\mt, \tys) \mapsto \aenv_\mt, (\lab_\mf, \tys) \mapsto
        \aenv_\mf \}, \emp) &
        \text{where} \;
        \aenv_\mt = \refine(\aenv, \expr, \true) \wedge
        \aenv_\mf = \refine(\aenv, \expr, \false)\\

        \asemi{\kwreturn \; \expr}
        (\lab, \tys, \aenv, \aelem) &=&
        (\{ (\lab_r, \tys_r) \mapsto \aenv_r \mid (\lab_r, \tys_r, \x) \in
        \retp(\getfunc(\lab), \tys)\}, \emp) &
        \text{where} \;
        \aelem = (\rmap, \retp) \wedge
        \aenv_r = \rmap(\lab_r, \tys_r)[\x \mapsto \aty] \wedge
        \aty = \aseme{\expr}(\aenv)\\

        \asemi{\x = \expr}
        (\lab, \tys, \aenv, \_) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv_\x \}, \emp) &
        \text{where} \;
        \aenv_\x = \aenv[\x \mapsto \aty] \wedge
        \aty = \aseme{\expr}(\aenv)\\

        \asemi{\refer \kwsl \expr_0 \kwsr = \expr_1}
        (\lab, \tys, \aenv, \_) &=&
        (\{ (\getnext(\lab), \tys) \mapsto \aenv \}, \emp)
      \end{array}
    \]
    \caption{Instructions with $\asemi{\inst}: (\labset \times \tyset^* \times
    \aenvset \times \adom) \rightarrow \adom$}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\textwidth}
    \[
      \asemr{\x}(\aenv) = \aenv(\x)
      \qquad
      \asemr{\refer \kwsl \expr \kwsr}(\aenv) = \{ \ty[\str] \mid \ty \in
      \asemr{\refer}(\aenv) \wedge \str \in \aseme{\expr}(\aenv) \}
    \]
    \caption{References with $\asemr{\refer}: \aenvset \rightarrow \atyset$}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \[
      \begin{array}{r@{~}c@{~}lr@{~}c@{~}lr@{~}c@{~}l}
        \aseme{\tname \; \kwcl \fields \kwcr}(\aenv)
        &=& \multicolumn{7}{@{}l}{
          \left\{
            \begin{array}{ll}
              \{ \tnormal(\ty) \mid \ty \in \aseme{\fields(\code{Value})}(\aenv)
              \} & \text{if} \; \tname = \text{Completion} \wedge
              \fields(\code{Type}) = \nconst{normal}\\

              \{ \tabrupt \} & \text{if} \; \tname = \text{Completion} \wedge
              \fields(\code{Type}) \neq \nconst{normal}\\

              \{ \tname \} & \text{otherwise}\\
            \end{array}
          \right.
          \quad \aseme{\prim}(\aenv) =
          \left\{
            \begin{array}{ll}
              \{ \tnum \} & \text{if} \; \prim = \num\\
              \{ \tbigint \} & \text{if} \; \prim = \bigint\\
              \{ \tsymb \} & \text{if} \; \prim = \symb\\
              \{ \prim \} & \text{otherwise}\\
            \end{array}
          \right.
        }

        \\

        \aseme{\const}(\aenv) &=& \const &

        \aseme{\clist{}}(\aenv) &=& \clist{} &

        \aseme{\clist{\expr_0, \cdots, \expr_n}}(\aenv)
        &=& \{ \clist{\ty} \mid \ty \in \bigjoin_{0 \leq i
        \leq n} \aseme{\expr_i}(\aenv) \}

        \\

        \aseme{\expr: \ty}(\aenv)
        &=& \{ \ty' \subtype \ty \mid \ty' \in
        \aseme{\expr}(\aenv) \} &

        \aseme{\refer \kwexists}(\aenv)
        &=& \{ \ty \neq \tabsent \mid \ty \in
        \aseme{\expr}(\aenv) \} &

        \aseme{\refer}(\aenv)
        &=& \asemr{\refer}(\aenv) \setminus \{ \tabsent \}

        \\

        \aseme{\uop \; \expr}(\aenv)
        &=& \{ \abs{\uop} \ty \mid \ty \in
        \aseme{\expr}(\aenv) \} &

        \aseme{\expr_0 \bop \expr_1}(\aenv)
        &=& \multicolumn{4}{@{}l}{\{ \ty_0 \abs{\bop} \ty_1 \mid \ty_0 \in
        \aseme{\expr_0}(\aenv) \wedge \ty_1 \in \aseme{\expr_1}(\aenv) \}}

        \\
      \end{array}
    \]
    \caption{Expressions with $\aseme{\expr}: \aenvset \rightarrow \atyset$}
  \end{subfigure}
  \caption{Abstract semantics of the modified $\ires$ for a program $\prog =
  (\getfunc, \getlab, \getnext)$.}
  \vspace*{-1.5em}
  \label{fig:abs-sem}
\end{figure*}

We design a type analysis for the modified $\ires$ based on abstract
interpretation framework~\cite{ai1977, ai1992} with analysis
sensitivity~\cite{sens-toplas}.  We first extend types $\tyset$ as follows:
\[
  \tyset \ni \ty ::=
  \cdots \mid
  \func \mid
  \const \mid
  \tnormal(\ty) \mid
  \tabrupt \mid
  \bool \mid
  \str \mid
  \tabsent
\]
There are three different reasons why we extend types.  First, the original
types $\tyset$ do not cover all types of values thus we add functions and
constants.  Second, we add more specific types to increase the precision of type
analysis.  A completion record is a \textit{normal completion} when its Type
field is a constant $\nconst{normal}$ or an \textit{abrupt completion}
otherwise.  We add $\tnormal(\ty)$ and $\tabrupt$ for normal and abrupt
completions.  Moreover, we also add Boolean values $\bool$, and String values
$\str$ to precisely handle the control flows of branches and field accesses,
respectively.  Third, we add the absent type $\tabsent$ to represent the
existence of variables.  Using such extended types, we define abstract states
with flow-sensitivity and type-sensitivity for parameters as follows:
\[
  \begin{array}{lr@{~}c@{~}c@{~}r@{~}l}
    \text{Abstract States}
    &\aelem&\in&\adom&=& \rmapset \times \retpset\\

    \text{Result Maps}
    &\rmap&\in&\rmapset&=& \labset \times \tyset^* \rightarrow \aenvset\\

    \text{Return Points}
    &\retp&\in&\retpset&=& \funcset \times \tyset^*
    \rightarrow \partsof{\labset \times \tyset^* \times \varset} \\

    \text{Abstract Environments}
    &\aenv&\in&\aenvset&=&\varset \rightarrow \atyset\\

    \text{Abstract Types}
    &\aty&\in&\atyset&=&\partsof{\tyset}\\
  \end{array}
\]
An abstract state $\aelem \in \adom$ is pair of a return map and a return point
mapping.  A result map $\rmap \in \rmapset$ represents an abstract environment
for each flow-sensitive and type-sensitive view, and a return point mapping
$\retp \in \retpset$ represents possible return points of each function with
a type-sensitive context.  An abstract environment $\aenv \in \aenvset$
represents possible types of each variable, and $\aenv(\x) = \{ \tabsent \}$
when $\x$ is not defined in $\aenv$.  An abstract type $\aty \in \atyset$ is a
set of extended types.

We define the abstract semantics of the modified $\ires$ as described in
Figure~\ref{fig:abs-sem} for a program $\prog = (\getfunc, \getlab,
\getnext)$.

### 2021-04-12 paper/ase21/overview.tex
[ASE21] Figure 5 - abstract semantics

\paragraph{Syntax and Semantics} ECMAScript uses a variant of extended BNF
notations to describe JavaScript syntax.  $\jiset$ synthesizes AST structures
and JavaScript parsers from its lexical and syntactic productions.  Our tool
utilizes only AST structures to define AST types and to access corresponding
syntax-directed algorithms.  For JavaScript semantics, ECMAScript contains
abstract algorithms written in a structured natural language.  $\jiset$ compiles
them to functions of $\ires$, an untyped intermediate representation for
ECMAScript.  We modify $\ires$ to make it more suitable for type analysis, and
formally define its syntax and semantics in Section~\ref{sec:ires}.

\paragraph{Types} In ECMAScript, types denote not only JavaScript types but also
specification types for ASTs, records, and lists.  In abstract algorithms, ASTs
are also values thus they can be stored in the variables or passed as arguments.
For such ASTs, we use their production names as their types and automatically
link corresponding syntax-directed algorithms to them as their fields.
ECMAScript also supports various record types and their fields and possible
values are defined in the corresponding tables.  For example, fields of the
``Realm'' record type are defined in ``Table 23: Realm Record Fields'' in
ECMAScript.  The latests version of ECMAScriptThe contains tens of tables for
record types and they are backward compatible in general.  Thus, we manually
model them to use them in type analysis.  For lists, we define the empty list
type $\clist{}$ and parametric list types $\clist{\tau}$.

>>>>

\paragraph{Syntax and Semantics} ECMAScript uses a variant of extended BNF
notations to describe JavaScript syntax.  $\jiset$ synthesizes AST structures
and JavaScript parsers from its lexical and syntactic productions.  Our tool
utilizes only AST structures to define their types and to access corresponding
syntax-directed algorithms.  For JavaScript semantics, ECMAScript contains
abstract algorithms written in a structured natural language.  $\jiset$ compiles
them to functions of $\ires$, an untyped intermediate representation for
ECMAScript.  We modify $\ires$ to make it more suitable for type analysis, and
formally define its syntax and semantics in Section~\ref{sec:ires}.

\paragraph{Types} In ECMAScript, types denote not only JavaScript types but also
specification types for ASTs, records, and lists.  In abstract algorithms, ASTs
are also values thus they can be stored in the variables or passed as arguments.
For such ASTs, we use their production names as their types and automatically
link corresponding syntax-directed algorithms to them as their fields.
ECMAScript also supports various record types and their fields and possible
values are defined in the corresponding tables.  For example, ``Table 9:
Completion Record Fields'' in the latest ECMAScript describes fields of the
completion records.  The latest version contains tens of tables for record types
but they are backward compatible in general.  Thus, we manually model fields of
record types based on the latest version to use them in type analysis.  For
lists, we define the empty list type $\clist{}$ and parametric list types
$\clist{\tau}$.
